Possible Inconsistency in Maximum Likelihood Calculation of Ancestral States

### 2017-06-23 joint_inf.tex
A pass through site split formulation.

\documentclass[a4paper]{article}

@@

%% Sets page size and margins
\usepackage[a4paper,top=3cm,bottom=2cm,left=3cm,right=3cm,marginparwidth=1.75cm]{geometry}

@@

For the $i$th factor of \eqref{eq:full_likelihood},
$$
P(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t) = P(\alignmentColumnRV=\alignmentColumn_i \mid \tau, t) \cdot P(\ancestralStateColumnRV=\ancestralStateColumn_i \mid \alignmentColumnRV=\alignmentColumn_i, \tau, t).
$$
As a consequence of assuming a binary symmetric model, taking complements yields
\begin{align*}
    2\cdot P(\alignmentColumnRV=\alignmentColumn_i \mid \tau, t) &= P(\siteSplitRV=\patternToSplit(\alignmentColumn_i) \mid \tau, t) \\
                                                                 &= P(\siteSplitRV=\siteSplit_j \mid \tau, t)
\end{align*}
for some $j$.
For the other term, we define
%EM Need a definition for \ancestralToSplit here.
$$
\ancestralToSplit:\alphabet^\nAncestralStateRows\times\alphabet^\nSiteRows\rightarrow\mathcal{P}(\ancestralSplitSet)
$$
as a mapping of ancestral states to ancestral state splits.
In this case, the ancestral states we consider depend on whether their corresponding tip labels had their complements taken or not: if we exclude a set of tip labels $\alignmentColumn$ from the set of site patterns by taking its complement, we let $\ancestralToSplit(\alignmentColumn, \ancestralStateColumn)$ be the complement of $\ancestralStateColumn$; otherwise $\ancestralStateColumn$ is unchanged.
We define
$$
\ancestralSplitRV := \ancestralToSplit(\alignmentColumnRV, \ancestralStateColumnRV)
$$
for an ancestral state--valued random variable $\ancestralStateColumnRV$ such that
\begin{align*}
    P(\ancestralStateColumnRV=\ancestralStateColumn_i \mid \alignmentColumnRV=\alignmentColumn_i, \tau, t) &= P(\ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \siteSplitRV=\patternToSplit(\alignmentColumn_i), \tau, t) \\
                                                                                                           &= P(\ancestralSplitRV=\ancestralSplit_k \mid \siteSplitRV=\siteSplit_j, \tau, t)
\end{align*}
for some $j$ and $k$.
Then, the $i$th factor of \eqref{eq:full_likelihood} has the property
$$
P(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t) \propto P(\siteSplitRV=\siteSplit_j, \tau, t) \cdot P(\ancestralSplitRV=\ancestralSplit_k \mid \siteSplitRV=\siteSplit_j, \tau, t).
$$

Taking the maximum across internal states yields
\begin{align*}
    \max_{\ancestralSplit_k\in\mathcal{P}(\ancestralSplitSet)} \ P(\siteSplitRV=\siteSplit_j \mid  \tau, t)\cdot P(\ancestralSplitRV=\ancestralSplit_k  \mid \siteSplitRV=\siteSplit_j, \tau, t) &=
P(\siteSplitRV=\siteSplit_j \mid  \tau, t)\cdot\max_{\ancestralSplit_k\in\mathcal{P}(\ancestralSplitSet)} \ P(\ancestralSplitRV=\ancestralSplit_k  \mid \siteSplitRV=\siteSplit_j, \tau, t).
\end{align*}
Given $(\tau, t)$, there exists an ordered list of sets $\fullAncestralSplitPartitions(\tau, t)=(\ancestralSplitPartition_1(\tau, t),\ldots,\ancestralSplitPartition_\nSiteSplits(\tau, t))$ such that any element $\xi_j$ of the $j$th component $\ancestralSplitPartition_j(\tau, t)$ satisfies
\begin{align*}
\max_{\ancestralSplit_k\in\mathcal{P}(\ancestralSplitSet)} \ P(\ancestralSplitRV=\ancestralSplit_k \mid \siteSplitRV=\siteSplit_j, \tau, t) &= P(\ancestralSplitRV = \xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t).
\end{align*}
Let $\xi_j$ be such a choice for each $1 \leq j \leq q$ in the following.
Then, the likelihood in \eqref{eq:profile_likelihood} written as a product over site patterns as opposed to sites is
\begin{align}
L_\nCols'(\tau, t; \fullAlignment) &= \max_{\fullAncestralStates} \ L_\nCols(\tau, t; \fullAlignment, \fullAncestralStates) \\
                             &= \prod_{i=1}^{\nCols} \ \max_{\ancestralStateColumn_i} \ P(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t) \\
%EM My suggestion is to use \psi(y_i) until you switch to a product indexed by the j's.
                             &\propto \prod_{i=1}^{\nCols} \ \max_{\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i)\in\mathcal{P}(\ancestralSplitSet)} \ P(\siteSplitRV=\patternToSplit(\alignmentColumn_i) \mid \tau, t) \cdot P(\ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \siteSplitRV=\patternToSplit(\alignmentColumn_i), \tau, t) \\
                             &= \prod_{i=1}^{\nCols} \ P(\siteSplitRV=\patternToSplit(\alignmentColumn_i) \mid \tau, t) \cdot \max_{\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i)\in\mathcal{P}(\ancestralSplitSet)} P(\ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \siteSplitRV=\patternToSplit(\alignmentColumn_i), \tau, t) \\
                             &= \prod_{j=1}^{\nSiteSplits} \ \left[P(\siteSplitRV=\siteSplit_j \mid \tau, t)\cdot P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t)\right] ^{\nCols_j(\fullAlignment)} \label{eq:site_pattern_likelihood}
\end{align}
where $\nCols_j(\fullAlignment)$ is the number of columns in $\fullAlignment$ where the site split $\siteSplit_j$ or its complement appears.

\paragraph{Example}
%EM My long term suggestion is for you to put this in example in the main text and the above in the appendix. (But we should probably check with Vladimir.)
\begin{figure}
    \centering
    \includegraphics[width=.45\textwidth]{unrooted_four_taxa}
    \caption{Four taxon tree}
\label{fig:four-taxa-tree}
\end{figure}

@@

We are interested in the properties of the objective function being maximized in \eqref{eq:site_pattern_likelihood} as data becomes plentiful.
Let
$$
L_\nCols''(\tau, t; \fullAlignment, \fullAncestralSplitPartitions(\tau, t)) = \prod_{j=1}^{\nSiteSplits} \ \left[P(\siteSplitRV=\siteSplit_j \mid \tau, t) \cdot P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t)\right] ^{\nCols_j(\fullAlignment)}
$$
be the final product in \eqref{eq:site_pattern_likelihood}.
Assume our $\nCols$ observations were generated from a model with parameters $(\tau^*, t^*)$.
We have
\begin{align}
    \frac{1}{\nCols} \log L_\nCols''(\tau, t; \fullAlignment, \fullAncestralSplitPartitions(\tau,t))
        &= \sum_{j=1}^\nSiteSplits \frac{\nCols_j(\fullAlignment)}{\nCols}\cdot  \log P(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_j \mid \tau, t) \\
        &= \sum_{j=1}^\nSiteSplits \frac{\nCols_j(\fullAlignment)}{\nCols}\cdot [\log P(\siteSplitRV=\siteSplit_j \mid \tau, t) +
            \log P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j , \tau, t)]
\end{align}
so that, in the limit of infinite data,
\begin{equation}
    \frac{1}{\nCols} \log L_\nCols''(\tau, t; \fullAlignment, \fullAncestralSplitPartitions(\tau, t)) \rightarrow \sum_{j=1}^\nSiteSplits P(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot [\log P(\siteSplitRV=\siteSplit_j \mid \tau, t) + \log P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j , \tau, t)]. \label{eq:site_pattern_profile_likelihood_mean}
\end{equation}
Define the divergence quantity
$$
\shannonDivergence_{\tau^*,t^*}(\tau,t) = \sum_{j=1}^\nSiteSplits P(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)\cdot\log P(\siteSplitRV=\siteSplit_j \mid \tau, t),
$$
and the partial log likelihood
$$
\tilde{\ell}_{\tau^*,t^*}(\tau, t; \fullAncestralSplitPartitions(\tau,t)) = \sum_{j=1}^\nSiteSplits P(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)\cdot\log P(\ancestralSplitRV=\xi_j \mid \siteSplit_j, \tau, t)
$$
so that \eqref{eq:site_pattern_profile_likelihood_mean} becomes
\begin{equation}
    \label{eq:log_likelihood_simplified}
    \ell_{\tau^*,t^*}(\tau, t; \fullAncestralSplitPartitions(\tau,t)) = \shannonDivergence_{\tau^*,t^*}(\tau,t) + \tilde{\ell}_{\tau^*,t^*}(\tau, t; \fullAncestralSplitPartitions(\tau,t)).
\end{equation}
For data generated from $(\tau^*, t^*)$, we have an inconsistency if
\begin{equation}
\label{eq:inconsistency_inequality}
\ell_{\tau^*,t^*}(\tau', \hat{t}_1; \fullAncestralSplitPartitions(\tau', \hat{t}_1)) > \ell_{\tau^*,t^*}(\tau^*, \hat{t}_2; \fullAncestralSplitPartitions(\tau^*, \hat{t}_2))
\end{equation}
for $\tau'\neq\tau^*$ with $\{\hat{t}_1,\fullAncestralSplitPartitions(\tau', \hat{t}_1)\}$ and $\{\hat{t}_2,\fullAncestralSplitPartitions(\tau^*, \hat{t}_2)\}$ estimated by maximizing $\tilde{\ell}_{\tau^*,t^*}$ with respect to the branch lengths and the internal states for $\tau = \tau'$ and $\tau^*$, respectively.
By Gibbs' inequality, $\shannonDivergence_{\tau^*,t^*}(\tau,t)$ is maximized when $(\tau,t) = (\tau^*,t^*)$, so any example of inconsistency must overcome that difference with substantially greater partial likelihoods on the non-generating topology.
In the next section, we show that such examples exist.

@@

Define the Farris topology $\tau_1$ and the Felsenstein topology $\tau_2$ as in Figure~\ref{fig:farris-fels-top}.
Call $t=\{x',y',w'\}$ and $t^*=\{x,y,y\}$, i.e., $t^*$ constrains the bottom three branches to share the same parameter, the classical construction of this topology.

We parameterize these trees \textbf{not with branch lengths}, but, rather, such that the probability of a substitution on a branch with parameter $x'$ is $p_x'=(1-x')/2$.
Here, $\{p_x',p_y',p_w'\}\in[0,1/2]^3$ and $\{x',y',w'\}\in[0,1]^3$.
This parametrization has been called edge ``fidelity'' as it quantifies the fidelity of transmission of the ancestral state along an edge \cite{Matsen2007-jq}.
Fidelities have useful algebraic properties, such as the ability to write generating probabilities using the Hadamard transform.

%das: did this on paper, but can write it out.
\textbf{Justification for estimand:} Consider the Farris topology with arbitrary branch lengths, i.e., $\tilde{t}=\{x_1,y_1,x_2,y_2,w\}$ counting from the top left.
The log likelihood $\ell_{\tau_1,t^*}(\tau_1, \tilde{t}; \fullAncestralSplitPartitions(\tau_1,\tilde{t}))$ is a function of $\{x_1x_2, x_1+x_2, y_1y_2, y_1+y_2, w\}$ only.
By symmetry, exchanging $x_1$ and $x_2$ does not change the value of $\ell$, requiring $x_1=x_2$ at the maximum.
A similar property holds for $y_1$ and $y_2$.
The Felsenstein topology does not admit this property, but, since we are interested in a lower bound for this topology, we simplify the objective function by constraining $x_1=x_2$ and $y_1=y_2$ similarly.

@@

>>>>

\documentclass{article}

@@

@@

For the $i$th factor of \eqref{eq:full_likelihood},
$$
P(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t) = P(\alignmentColumnRV=\alignmentColumn_i \mid \tau, t) \cdot P(\ancestralStateColumnRV=\ancestralStateColumn_i \mid \alignmentColumnRV=\alignmentColumn_i, \tau, t).
$$
As a consequence of assuming a binary symmetric model, taking complements yields
\begin{align*}
    2\cdot P(\alignmentColumnRV=\alignmentColumn_i \mid \tau, t) &= P(\siteSplitRV=\patternToSplit(\alignmentColumn_i) \mid \tau, t) \\
                                                                 &= P(\siteSplitRV=\siteSplit_j \mid \tau, t)
\end{align*}
for some $j$.
For the other term, we define
$$
\ancestralToSplit:\alphabet^\nAncestralStateRows\times\alphabet^\nSiteRows\rightarrow\mathcal{P}(\ancestralSplitSet)
$$
as a mapping of ancestral states and tip states to ancestral state splits.
It is defined by if the tip states have their complements taken or not: if we exclude a set of tip labels $\alignmentColumn$ from the set of site patterns by taking its complement, we let $\ancestralToSplit(\alignmentColumn, \ancestralStateColumn)$ be the complement of $\ancestralStateColumn$; otherwise it is simply $\ancestralStateColumn$.
Note $\ancestralToSplit(\alignmentColumn_i, \cdot)$ is surjective for each $\alignmentColumn_i$.
We define
$$
\ancestralSplitRV := \ancestralToSplit(\alignmentColumnRV, \ancestralStateColumnRV)
$$
for a tip state--valued random variable $\alignmentColumnRV$ and an ancestral state--valued random variable $\ancestralStateColumnRV$.
By considering the case requiring complementing and that not requiring complementing separately,
\begin{align*}
    P(\ancestralStateColumnRV=\ancestralStateColumn_i \mid \alignmentColumnRV=\alignmentColumn_i, \tau, t) &= P(\ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \siteSplitRV=\patternToSplit(\alignmentColumn_i), \tau, t).
\end{align*}
Given $(\tau, t)$, there exists an ordered list of sets $\fullAncestralSplitPartitions(\tau, t)=(\ancestralSplitPartition_1(\tau, t),\ldots,\ancestralSplitPartition_\nSiteSplits(\tau, t))$ such that any element $\xi_j$ of the $j$th component $\ancestralSplitPartition_j(\tau, t)$ satisfies
\begin{align*}
\max_{\ancestralSplit_k\in\mathcal{P}(\ancestralSplitSet)} \ P(\ancestralSplitRV=\ancestralSplit_k \mid \siteSplitRV=\siteSplit_j, \tau, t) &= P(\ancestralSplitRV = \xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t).
\end{align*}
By the surjectivity of $\ancestralToSplit(\alignmentColumn_i, \cdot)$, we also have
\begin{align*}
\max_{\ancestralStateColumn_i} P(\ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \siteSplitRV=\patternToSplit(\alignmentColumn_i), \tau, t) &= P(\ancestralSplitRV = \xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t).
\end{align*}

Let $\xi_j$ be such a choice for each $1 \leq j \leq q$ in the following.
Then, the likelihood in \eqref{eq:profile_likelihood} written as a product over site patterns as opposed to sites is
\begin{align}
L_\nCols'(\tau, t; \fullAlignment) &= \max_{\fullAncestralStates} \ L_\nCols(\tau, t; \fullAlignment, \fullAncestralStates) \\
                             &= \prod_{i=1}^{\nCols} \ \max_{\ancestralStateColumn_i} \ P(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t) \\
                             &\propto \prod_{i=1}^{\nCols} \ \max_{\ancestralStateColumn_i} \ P(\siteSplitRV=\patternToSplit(\alignmentColumn_i) \mid \tau, t) \cdot P(\ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \siteSplitRV=\patternToSplit(\alignmentColumn_i), \tau, t) \\
                             &= \prod_{i=1}^{\nCols} \ P(\siteSplitRV=\patternToSplit(\alignmentColumn_i) \mid \tau, t) \cdot \max_{\ancestralStateColumn_i} P(\ancestralSplitRV=\ancestralToSplit(\alignmentColumn_i, \ancestralStateColumn_i) \mid \siteSplitRV=\patternToSplit(\alignmentColumn_i), \tau, t) \\
                             &= \prod_{j=1}^{\nSiteSplits} \ \left[P(\siteSplitRV=\siteSplit_j \mid \tau, t)\cdot P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t)\right] ^{\nCols_j(\fullAlignment)} \label{eq:site_pattern_likelihood}
\end{align}
where $\nCols_j(\fullAlignment)$ is the number of columns in $\fullAlignment$ where the site split $\siteSplit_j$ or its complement appears.

\paragraph{Example}
\begin{figure}
    \centering
    \includegraphics[width=.45\textwidth]{unrooted_four_taxa}
    \caption{Four taxon tree}
\label{fig:four-taxa-tree}
\end{figure}

@@

We are interested in the properties of the objective function being maximized in \eqref{eq:site_pattern_likelihood} as data becomes plentiful.
Let
$$
L_\nCols''(\tau, t; \fullAlignment, \fullAncestralSplitPartitions(\tau, t)) = \prod_{j=1}^{\nSiteSplits} \ \left[P(\siteSplitRV=\siteSplit_j \mid \tau, t) \cdot P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t)\right] ^{\nCols_j(\fullAlignment)}
$$
be the final product in \eqref{eq:site_pattern_likelihood}.
Assume our $\nCols$ observations were generated from a model with parameters $(\tau^*, t^*)$.
We have
\begin{align}
    \frac{1}{\nCols} \log L_\nCols''(\tau, t; \fullAlignment, \fullAncestralSplitPartitions(\tau,t))
        &= \sum_{j=1}^\nSiteSplits \frac{\nCols_j(\fullAlignment)}{\nCols}\cdot  \log P(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_j \mid \tau, t) \\
        &= \sum_{j=1}^\nSiteSplits \frac{\nCols_j(\fullAlignment)}{\nCols}\cdot [\log P(\siteSplitRV=\siteSplit_j \mid \tau, t) +
            \log P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j , \tau, t)]
\end{align}
so that, in the limit of infinite data,
\begin{equation}
    \frac{1}{\nCols} \log L_\nCols''(\tau, t; \fullAlignment, \fullAncestralSplitPartitions(\tau, t)) \rightarrow \sum_{j=1}^\nSiteSplits P(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot [\log P(\siteSplitRV=\siteSplit_j \mid \tau, t) + \log P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j , \tau, t)]. \label{eq:site_pattern_profile_likelihood_mean}
\end{equation}
Define the divergence quantity
$$
\shannonDivergence_{\tau^*,t^*}(\tau,t) = \sum_{j=1}^\nSiteSplits P(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)\cdot\log P(\siteSplitRV=\siteSplit_j \mid \tau, t),
$$
and the partial log likelihood
$$
\tilde{\ell}_{\tau^*,t^*}(\tau, t; \fullAncestralSplitPartitions(\tau,t)) = \sum_{j=1}^\nSiteSplits P(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)\cdot\log P(\ancestralSplitRV=\xi_j \mid \siteSplitRV = \siteSplit_j, \tau, t)
$$
so that \eqref{eq:site_pattern_profile_likelihood_mean} becomes
\begin{equation}
    \label{eq:log_likelihood_simplified}
    \ell_{\tau^*,t^*}(\tau, t; \fullAncestralSplitPartitions(\tau,t)) = \shannonDivergence_{\tau^*,t^*}(\tau,t) + \tilde{\ell}_{\tau^*,t^*}(\tau, t; \fullAncestralSplitPartitions(\tau,t)).
\end{equation}
For data generated from $(\tau^*, t^*)$, we have an inconsistency if
\begin{equation}
\label{eq:inconsistency_inequality}
\ell_{\tau^*,t^*}(\tau', \hat{t}_1; \fullAncestralSplitPartitions(\tau', \hat{t}_1)) > \ell_{\tau^*,t^*}(\tau^*, \hat{t}_2; \fullAncestralSplitPartitions(\tau^*, \hat{t}_2))
\end{equation}
for $\tau'\neq\tau^*$ with $\{\hat{t}_1,\fullAncestralSplitPartitions(\tau', \hat{t}_1)\}$ and $\{\hat{t}_2,\fullAncestralSplitPartitions(\tau^*, \hat{t}_2)\}$ estimated by maximizing $\tilde{\ell}_{\tau^*,t^*}$ with respect to the branch lengths and the internal states for $\tau = \tau'$ and $\tau^*$, respectively.
By Gibbs' inequality, $\shannonDivergence_{\tau^*,t^*}(\tau,t)$ is maximized when $(\tau,t) = (\tau^*,t^*)$, so any example of inconsistency must overcome that difference with substantially greater partial likelihoods on the non-generating topology.
In the next section, we show that such examples exist.

@@

Define the Farris topology $\tau_1$ and the Felsenstein topology $\tau_2$ as in Figure~\ref{fig:farris-fels-top}.
We parameterize the branches of these trees \textbf{not with branch lengths}, but with a parameter such that the probability of a substitution on a branch with parameter $\theta$ is $p_\theta = (1-\theta)/2$.
This parametrization has been called edge ``fidelity'' as it quantifies the fidelity of transmission of the ancestral state along an edge \cite{Matsen2007-jq}.
Fidelities have useful algebraic properties, such as the ability to write generating probabilities using the Hadamard transform.

Let $\{x',y',w'\}\in[0,1]^3$ be three fidelities and $\{p_x',p_y',p_w'\}\in[0,1/2]^3$ the corresponding substitution probabilities.
Call $t=\{x',y',w'\}$ and $t^*=\{x,y,y\}$, i.e., $t^*$ constrains the bottom three branches to share the same parameter, the classical construction of this topology.

%das: did this on paper, but can write it out.
\textbf{Justification for estimand:} Consider the Farris topology with arbitrary branch lengths, i.e., $\tilde{t}=\{x_1,y_1,x_2,y_2,w\}$ with edge ordering in the order of the taxa, then the internal branch last.
%EM Yeah, this next statement feels like it could use some justification, and or maybe you could skip to the following statement.
The log likelihood $\ell_{\tau_1,t^*}(\tau_1, \tilde{t}; \fullAncestralSplitPartitions(\tau_1,\tilde{t}))$ is a function of $\{x_1x_2, x_1+x_2, y_1y_2, y_1+y_2, w\}$ only.
By symmetry, exchanging $x_1$ and $x_2$ does not change the value of $\ell$, requiring $x_1=x_2$ at the maximum.
A similar property holds for $y_1$ and $y_2$.
The Felsenstein topology does not admit this property, but, since we are interested in a lower bound for this topology, we simplify the objective function by constraining $x_1=x_2$ and $y_1=y_2$ similarly.

@@

%EM I'd suggest p_\emptyset for this particular one, but perhaps that will throw off your spacing below.
